<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Data update}
-->

ENCODExplorer: A compilation of metadata from ENCODE
====================================================================
Audrey Lemacon, Louis Gendron, Charles Joly Beauparlant and Arnaud Droit.

This package and the underlying ENCODExplorer code are distributed under 
the Artistic license 2.0. You are free to use and redistribute this software. 

"The ENCODE (Encyclopedia of DNA Elements) Consortium is an international 
collaboration of research groups funded by the National Human Genome Research 
Institute (NHGRI). The goal of ENCODE is to build a comprehensive parts list of 
functional elements in the human genome, including elements that act at the 
protein and RNA levels, and regulatory elements that control cells and 
circumstances in which a gene is active"^[source : [ENCODE Projet Portal ](https://www.encodeproject.org/)] .

However, data retrieval and downloading can be really time-consuming using 
the current web portal. 

This package has been designed to facilitate data access by compiling the 
metadata associated with files, experiments and datasets. 

We first extract the ENCODE schema from its public github repository. 
We then identify the main entities and their relationships with each other to 
rebuild the ENCODE database into a data.table. 
We also developped a function which can extract the essential metadata into 
an R object to aid data exploration.
We implemented time-saving features to select ENCODE files by querying their 
metadata and download them.

The data.table can be regenerated at will to query an up-to-date ENCODE
database locally.

This vignette will introduce how to update the local copy of the ENCODE data.

### Loading the ENCODExplorer package

```{r libraryLoad}
suppressMessages(library(ENCODExplorer))
```

### Data update

Updating ENCODExplorer's local cache of ENCODE metadata relies on two functions 
from the ENCODExplorer package: 
* **prepare_ENCODEdb**, which fectches all raw metadata from ENCODE's website
using its REST API, and 
* **export_ENCODEdb_matrix**, which converts that metadata into an easy to 
parse data.table used by ENCODExplorer's query functions.

Thus, the first step to updating the local cache of ENCODE data is to call
**prepare_ENCODEdb**. By default, the results are stored in the current work 
directory to a file called "tables.RDA". Since fetching ENCODE metadata 
is a lengthy process, it is advisable to fetch this data only once, then to rely
on the cached data:

```{r tables, eval=FALSE}
# The path (relative or absolute) to the future database
database_filename = "new.encode.rda"
tables = prepare_ENCODEdb(database_filename)
```

Once the data has been fetched, it can be converted into a data.table equivalent 
to the default encode_df using **export_ENCODEdb_matrix**:

```{r, eval=FALSE}
# The path (relative or absolute) to the database downloaded using
# prepare_ENCODEdb.
database_filename <- "new.encode.rda"
new_encode_df <- export_ENCODEdb_matrix(database_filename)
```

The whole process will take several minutes, and will often need up to an hour
to complete.

### Using the updated data

You can use your new metadata by passing it to the queryEncode and downloadEncode 
functions using their *df* parameter:

```{r queryEncode, eval=FALSE}
query_results <- queryEncode(df = new_encode_df, assay = "switchgear",
                             target ="elavl1", file_format = "bed" , fixed = F)
downloadEncode(df = new_encode_df, file_acc = query_results$file_accession)
```

Be sure to use the same *df* data.table for both **queryEncode** and **downloadEncode**.

You can also use the generated data.table database for your own purposes. The
imputed database model is available in the [dedicated vignette](DBmodel.html)

### Updating the default package data

If you want to update ENCODExplorer's default ENCODE metadata so that you won't
have to call export_ENCODEdb_matrix with every new R session, you can replace
the tables.RDA file which came with ENCODExplorer. This can only be done if you
have write permissions for ENCODExplorer's install directory:

```{r replaceTables, eval=FALSE}
package_table = system.file("extdata", "tables.RDA", package="ENCODExplorer")
prepare_ENCODEdb(database_filename=package_table, overwrite=TRUE)
```

After this operation, ENCODExplorer will use the updated data whenever it
loads encode_df.